%% ============================================================
%% glossary.tex — Standalone English Glossary
%% (riusabile anche fuori dal manuale principale)
%% ============================================================

\chapter{Glossary of Historical and Game Terms}
\markboth{Glossary}{Glossary}
\begin{english}

\textit{This glossary collects all historical and game-specific terms
used in \textit{Vespri 1282}. Historical terms are marked (H);
game system terms are marked (G).}

\secsep

\begin{description}[leftmargin=2em, style=nextline, font=\cinzel\color{oro}]

\item[Action Roll (G)] The core mechanic of Forged in the Dark. Gather
  dice equal to your action rating, roll, take the highest: 6 = full
  success; 4--5 = partial success; 1--3 = failure.

\item[Angevin (H)] Of or relating to the House of Anjou. Charles of
  Anjou (r.~1266--1285) was the Angevin king of Sicily and Naples.
  In Sicilian usage, \textit{li Francisi} or \textit{li Provenzali}
  referred to all continental officials and soldiers regardless of
  precise origin.

\item[Aragonese (H)] Of or relating to the Crown of Aragon, a medieval
  Iberian kingdom. Peter III of Aragon claimed Sicily through his wife
  Constance, daughter of Manfred of Hohenstaufen.

\item[Attribute (G)] One of four character traits in \textit{Vespri 1282}:
  Heart, Mind, Body, Spirit. Each governs three actions.

\item[Bailli / Giustiziere (H)] Chief royal official of a province,
  responsible for justice, taxation, and military order. Jean de
  Saint-Rémy was the giustiziere of Palermo in 1282.

\item[Burden (G)] A structural weakness built into each playbook.
  Burdens create narrative complications specific to each character
  archetype.

\item[Cassaro (H)] The historic centre of Palermo, from Arabic
  \textit{al-Qasr} (the castle). Seat of the Norman Palace and
  Angevin administrative power.

\item[Conjunction Moment (G)] One of three structured moments during
  the session when the coordinating GM distributes shared events to
  all tables. See: The Rumour, The Crisis, The Bells.

\item[Controlled / Risky / Desperate (G)] The three positions in the
  FitD system, indicating the level of exposure to harm before a roll.

\item[De Saint-Rémy Clock (G)] A 6-segment hidden clock tracking Angevin
  awareness. When full: curfew, port closure, reinforced patrols, -1
  to all faction scores.

\item[Effect (G)] The scale of impact of a successful action:
  Limited, Standard, or Great.

\item[Emirate of Sicily (H)] The Islamic polity governing Sicily
  (c.~831--1072 CE). Its legacy was visible in Palermo's urban structure,
  agriculture, and cultural practice in 1282.

\item[Excommunication (H)] Formal exclusion from the Catholic Church.
  A major political weapon. Pope Martin IV excommunicated Michael VIII
  Palaeologus in 1281.

\item[Faction (G)] One of the four playing groups in \textit{Vespri 1282}:
  Conspirators, Barons, Clergy, People. Each faction plays at its own
  table and contributes to the shared outcome score.

\item[Faction Score (G)] A 0--4 numerical value tracking each faction's
  effectiveness during the session. The total (0--16) determines the
  final outcome.

\item[Fealty / Homage (H)] The sworn loyalty of a vassal to their lord,
  the foundation of feudal social organisation. The Barons' faction is
  defined by the tension between their fealty to Charles and their
  Sicilian identity.

\item[Forged in the Dark / FitD (G)] The game system used by
  \textit{Vespri 1282}, created by John Harper for \textit{Blades in
  the Dark} (2017). Focused on desperate characters making difficult
  choices under pressure.

\item[Fortune (G)] A mechanic allowing players to push past their
  character's limits by adding 1d6 to a roll at the cost of 2 Stress.

\item[Gift (G)] A special ability unique to each playbook, providing
  a narrative or mechanical advantage tied to the character's archetype.

\item[Hara / Harat (H)] Arabic term for a neighbourhood within a city.
  Palermo's quarter structure in 1282 reflected its Arabic urban heritage.

\item[Hohenstaufen (H)] The German imperial dynasty (also: Staufen,
  Swabian) ruling Sicily 1194--1268. Frederick II and Manfred were its
  greatest Sicilian representatives. Their fall creates the power vacuum
  at the heart of \textit{Vespri 1282}.

\item[In Medias Res (G)] The Conspirators' entry point: characters begin
  already inside the conspiracy, with no recruitment scene.

\item[Kalsa (H)] Historic quarter of Palermo near the port, from Arabic
  \textit{al-Khalisa} (the pure). Densely populated, Arab in heritage.

\item[Latin Empire of Constantinople (H)] The crusader state in
  Constantinople (1204--1261), whose restoration Charles of Anjou planned.
  Byzantine support for the conspiracy was motivated by the desire to
  block this plan.

\item[Limited / Standard / Great (G)] The three levels of Effect in the
  FitD system, determining how much a character's action changes the
  situation.

\item[Norman Palace / Palazzo dei Normanni (H)] Seat of Angevin power
  in Palermo. Built by Arab craftsmen for Norman rulers, adapted by
  successive dynasties. Today houses the Sicilian Regional Assembly.

\item[OpenHistoryMap / OHM (G/H)] An open geographic platform for
  historical data. The OHM appendix of this manual provides RDF/Turtle
  data for Palermo 1282 and other Sicilian centres, usable as a
  standalone historical resource.

\item[Playbook (G)] A character archetype in Forged in the Dark games,
  providing peak actions, special abilities (Gift), structural weaknesses
  (Burden), and an opening Bond. In \textit{Vespri 1282}, playbooks
  are historically grounded archetypes.

\item[Position (G)] See Controlled / Risky / Desperate.

\item[Provence (H)] Region of southern France, homeland of many Angevin
  officials. Sicilians used \textit{provenzali} (Provençals) alongside
  \textit{francesi} (French) for their rulers.

\item[Resistance Roll (G)] A roll to reduce or negate a consequence.
  Always costs 1 Stress regardless of result.

\item[rpg-schema.org (G)] An open ontology for roleplaying game data,
  used in the semantic appendix of this manual to formally describe
  the game's elements.

\item[Santo Spirito (H)] Norman church outside Palermo's walls. Site
  of the Easter Monday procession of 30 March 1282. The spark of
  the Vespers uprising. Also called \textit{chiesa dei Vespri}.

\item[Stress (G)] A resource tracking psychological and physical pressure.
  9 boxes per character. Filling the ninth box triggers a Trauma.

\item[Telegram Bot (G)] A simple coordination tool for the multi-table
  structure. GMs send faction scores and shared events via bot commands.

\item[Trauma (G)] A permanent psychological mark acquired when Stress
  fills. Types in \textit{Vespri 1282}: Distrust, Obsession, Guilt,
  Fear, Lost Faith. Four traumas: character is out of play.

\item[Vassal (H)] A person owing loyalty and service to a lord in
  exchange for protection and land. Feudal Sicily's vassal webs were
  disrupted and partially replaced by the Angevin conquest.

\item[Vespers / Vespro (H)] The evening canonical hour, typically at
  sunset. The uprising of 30 March 1282 began during or just after
  vespers at Santo Spirito, giving it its name. \textit{I Vespri
  Siciliani} = the Sicilian Vespers.

\end{description}

\end{english}
