%% en/sections/04_playbook.tex — Playbooks EN (full)

\chapter{The Playbooks}
\markboth{The Playbooks}{The Playbooks}

\lettrine[lines=3]{E}{ach} character in \textit{Vespri 1282} is defined
by a \textbf{playbook}: a historical archetype fixing their narrative
role, peak actions, and central tension. Each playbook provides:
\textbf{peak actions} (rating 2), \textbf{secondary actions} (rating 1),
a \textbf{Gift} (special ability), a \textbf{Burden} (structural
weakness), and an opening \textbf{Bond} with another PC in the same
faction.

\secsep

%% ============================================================
\section{The Conspirators — Playbooks}

\subsection{The Messenger}
{\cinzel\footnotesize\color{grigionote} Ruggero di Amato}\\
\textit{You carry words that can change the world. Or get you killed.}

\textbf{Peak actions:} Manoeuvre, Deceive\\
\textbf{Secondary actions:} Study, Act, Persuade, Resist

\textbf{Gift --- Multiple Faces:} You have at least three ready false
identities. When you use one, your Deceive roll is always at least
Controlled position, unless they have already seen you in that face.

\textbf{Burden --- The Chain:} Every message you carry binds you to
its sender. If captured, the GM may call for a Desperate Resist roll
to avoid revealing the source. If you fail, the faction loses 1 point.

\textbf{Opening Bond:} The Conspirator knows your real name.
You are the only one who does.

\filetto

\subsection{The Conspirator}
{\cinzel\footnotesize\color{grigionote} Gualtiero de Caltagirone}\\
\textit{You know the plan. Or at least, your piece of it.}

\textbf{Peak actions:} Plot, Study\\
\textbf{Secondary actions:} Deceive, Persuade, Command, Resist

\textbf{Gift --- Compartmentalisation:} Once per session, you may
declare that something you are doing was ``part of the plan'' from
the beginning, even if it was not previously stated. This does not
change the dice — it changes the narrative.

\textbf{Burden --- The Weight of Knowing:} You know things the others
do not. When another faction PC fails a Study roll to understand the
situation, you may choose to reveal the information — but suffer
1 Stress for doing so.

\textbf{Opening Bond:} The Messenger carries your words.
You hope they are careful enough.

\filetto

\subsection{The Spy}
{\cinzel\footnotesize\color{grigionote} Margherita di Brindisi}\\
\textit{You work inside the Angevin system. You have for too long.}

\textbf{Peak actions:} Deceive, Study\\
\textbf{Secondary actions:} Manoeuvre, Act, Persuade, Resist

\textbf{Gift --- Inside the Palace:} You have physical access to
spaces others cannot reach: the Norman Palace, the justiciars' offices,
the military quarters — as long as you maintain your cover.

\textbf{Burden --- Contamination:} You have lived among the French
too long. When you act against an Angevin official with whom you have
a personal relationship, roll Resist or suffer 1 Stress from hesitation.

\textbf{Opening Bond:} The Conspirator recruited you three years ago.
You have not yet forgiven yourself for accepting.

\filetto

\subsection{The Merchant}
{\cinzel\footnotesize\color{grigionote} Bartolomeo Lancia}\\
\textit{Money has no flag. But you do.}

\textbf{Peak actions:} Persuade, Plot\\
\textbf{Secondary actions:} Study, Deceive, Command, Manoeuvre

\textbf{Gift --- Trade Network:} You have contacts in every Sicilian
port and beyond. You may request information, goods, or safe passage
without a roll — but each request has a narrative cost.

\textbf{Burden --- The Open Account:} You funded part of the conspiracy
with your own money. If the uprising fails, you are ruined. Whenever
a plan seems to waver, roll Resist or act selfishly.

\textbf{Opening Bond:} The Spy provided you with false documents that
saved your life. You have not yet repaid them.

\filetto

\subsection{The Deserter}
{\cinzel\footnotesize\color{grigionote} Drogo de Metz}\\
\textit{You fought for Charles of Anjou. Then you saw what you were defending.}

\textbf{Peak actions:} Fight, Resist\\
\textbf{Secondary actions:} Act, Manoeuvre, Command, Defy

\textbf{Gift --- Military Discipline:} When you protect another PC from
physical harm, the harm is reduced by 1 box. You may do this even in
Desperate position without a roll.

\textbf{Burden --- The Known Face:} Angevin soldiers in Palermo
recognise you. Whenever you move in patrolled areas, the GM may declare
that someone has spotted you — unless you took specific precautions.

\textbf{Opening Bond:} The Messenger convinced you to switch sides.
You were not difficult to convince.

\filetto

\subsection{The Noble in Exile}
{\cinzel\footnotesize\color{grigionote} Tancredi di Chiaramonte}\\
\textit{Descendant of a family Frederick loved. Charles hates.}

\textbf{Peak actions:} Command, Persuade\\
\textbf{Secondary actions:} Study, Plot, Resist, Defy

\textbf{Gift --- The Ancient Name:} Your surname opens doors among
Sicilian barons. When you present yourself openly, barons cannot
ignore you — even if they do not trust you.

\textbf{Burden --- The Target:} You are on the Angevin surveillance
list. Every time you step into the open, the GM may raise the Angevin
alert level by one degree.

\textbf{Opening Bond:} The Conspirator knows your brother, still in
Barcelona. The news is not always good.

\filetto

\subsection{The Forger}
{\cinzel\footnotesize\color{grigionote} Niccolò the Cretan}\\
\textit{Documents, seals, identities. Everything that does not exist, you create.}

\textbf{Peak actions:} Deceive, Act\\
\textbf{Secondary actions:} Study, Plot, Manoeuvre, Resist

\textbf{Gift --- The Right Document:} Once per scene, you may declare
you have a document that resolves a bureaucratic or access problem.
No roll required — but it only holds until examined by an expert.

\textbf{Burden --- The Trace:} Every document you produce carries a
stylistic signature. After the third document used in the session,
the GM may declare that someone has noticed the similarity.

\textbf{Opening Bond:} The Merchant is your main client.
They know too much about you to be just a client.

\filetto

\subsection{The Engineer}
{\cinzel\footnotesize\color{grigionote} Greta degli Ingegni}\\
\textit{You build what the network cannot buy. Every tool of
yours is either a weapon or an alibi.}

\textbf{Peak actions:} Study, Act\\
\textbf{Secondary actions:} Plot, Manoeuvre, Persuade, Resist

\textbf{Background:} Greta grew up in a Cassaro goldsmith's shop
that also repaired Norman clocks and Arab mechanisms surviving
from Roger II's day. When her father died, she took over the
commissions --- and patrons discovered her hand was steadier than
his. The Merchant approached her two years ago for something
more delicate than a court alarm clock: sealed tubes that break
if forced, mirrors mounted on pivots that send signals across
miles, locks that open only to a sequence of taps. What she
makes for the conspiracy appears in no register. What she makes
for the Angevin court is her alibi.

\textbf{Gift --- The Right Object:} Once per scene, you may declare
you brought (or are quickly building) a small mechanical device
that solves a specific problem: a lockpick, a signalling mirror,
a cipher tube, a lock decoy, a timed fuse. You and the GM agree
what it does. No roll: it works the first time. It breaks or
is consumed in use.

\textbf{Burden --- The Master's Signature:} Every piece you make
carries traces of your hand. Keep count: after the third device
you leave behind or that falls into Angevin hands, the GM may
declare someone is connecting the pieces. The network loses
1 point.

\textbf{Opening Bond:} The Merchant funds you. They know your tools
are worth gold --- they suspect they are worth far more, and that
worries them.

\secsep

%% ============================================================
\section{The Barons — Playbooks}

\subsection{The Baron}
{\cinzel\footnotesize\color{grigionote} Simone di Mazzarino}\\
\textit{You have land, men, and obligations to both. And to your son.}

\textbf{Peak actions:} Command, Persuade\\
\textbf{Secondary actions:} Study, Plot, Defy, Resist

\textbf{Gift --- The Given Word:} When you make an agreement with
another character and declare it explicitly, both gain +1 die on rolls
relating to that agreement while it holds. If one breaks it, the other
gains a bonus die against them.

\textbf{Burden --- The Weight of the Fief:} Your decisions affect
more than you. Whenever you choose a risky action, the GM may ask:
\textit{and your men? And your family?} You must answer before rolling.

\textbf{Opening Bond:} The Knight is loyal to you. You hope that
loyalty survives the choices you are about to make.

\filetto

\subsection{The Knight}
{\cinzel\footnotesize\color{grigionote} Riccardo Bonello}\\
\textit{You swore loyalty. To whom, exactly, has become complicated.}

\textbf{Peak actions:} Fight, Defy\\
\textbf{Secondary actions:} Command, Resist, Act, Manoeuvre

\textbf{Gift --- Training:} When in combat with at least one ally
nearby, you are always in Risky position (not Desperate) unless
alone and surrounded.

\textbf{Burden --- The Old Oath:} You swore to Charles of Anjou (or
a vassal) when you had no choice. That oath still exists. Every time
you act against Angevin interests, roll Resist or suffer 1 Stress.

\textbf{Opening Bond:} The Baron protects you.
You are not sure you deserve it.

\filetto

\subsection{The Chatelaine}
{\cinzel\footnotesize\color{grigionote} Elisabetta di Paternò}\\
\textit{Your husband is dead. The castle is yours. Charles's men think they can deal with a woman.}

\textbf{Peak actions:} Command, Plot\\
\textbf{Secondary actions:} Persuade, Study, Resist, Deceive

\textbf{Gift --- The Underestimation:} Once per scene, you may declare
that an opponent has underestimated you. Your next roll against that
opponent is in Controlled position, even if circumstances would suggest
otherwise.

\textbf{Burden --- Visibility:} A woman who commands is visible.
Every Manoeuvre or Deceive roll in French-controlled areas starts at
Risky, never Controlled.

\textbf{Opening Bond:} The Notary holds your documents.
Documents that should not exist.

\filetto

\subsection{The Notary}
{\cinzel\footnotesize\color{grigionote} Matteo di Termini}\\
\textit{The law is a tool. The question is who wields it.}

\textbf{Peak actions:} Study, Plot\\
\textbf{Secondary actions:} Persuade, Deceive, Command, Resist

\textbf{Gift --- Living Archive:} You know the legal precedents,
treaties, and royal concessions of Norman and Hohenstaufen Sicily.
You may use Study to find a legal basis for almost any action —
which does not make it safe, but makes it justifiable.

\textbf{Burden --- The Uncomfortable Document:} You hold a document
that could shift the balance of power — and endangers you merely by
possession. The GM knows what it says. You have not yet read it to
the end.

\textbf{Opening Bond:} The Chatelaine trusts you.
You fear that trust is misplaced.

\filetto

\subsection{The Son}
{\cinzel\footnotesize\color{grigionote} Goffredo di Mazzarino}\\
\textit{Your father negotiates. You want to act.}

\textbf{Peak actions:} Defy, Act\\
\textbf{Secondary actions:} Fight, Resist, Incite, Manoeuvre

\textbf{Gift --- Impetus:} When you act first in a crisis situation,
without waiting for instructions, gain +1d. If your action succeeds,
add +1 to the faction score.

\textbf{Burden --- Inexperience:} Your starting position in any
situation requiring negotiation or diplomacy is always Risky, never
Controlled. You are still learning.

\textbf{Opening Bond:} The Knight taught you to fight.
They have not yet taught you when not to.

\filetto

\subsection{The Physician}
{\cinzel\footnotesize\color{grigionote} Master Filippo Carafa da Salerno}\\
\textit{You treat everyone, even the French. Especially the French — they know things.}

\textbf{Peak actions:} Study, Persuade\\
\textbf{Secondary actions:} Resist, Deceive, Act, Invoke

\textbf{Background:} Trained at the School of Salerno, Filippo Carafa
arrived in Palermo with an impeccable reputation — and certain interests
the School would not approve. Alongside Galenic medicine, he has studied
Arabic treatises that circulate only in restricted circles: herbs,
infusions, preparations that heal where standard remedies fail and bring
sleep where pain does not relent. He found in Yusuf ibn Khalaf al-Balarmì
a precious interlocutor — the man knows things no Salernitan physician
does — and in Maimone di Palermo a reliable supplier of materials
unavailable in official markets. The Baron knows about the School.
He does not know about the rest.

\textbf{Gift --- The Patient's Trust:} When you have treated someone,
they are in your debt. You may request information or a small favour
from anyone you have treated during the session, without a roll —
but only once per person.

\textbf{Burden --- The Hippocratic Oath:} You cannot refuse to treat
someone in need, not even an Angevin soldier. If you do, suffer
2 immediate Stress.

\textbf{Opening Bonds:} The Baron pays you. But you also treat the
poor of the Albergheria, free, at night. \textit{Cross-faction bonds
(optional):} Yusuf ibn Khalaf al-Balarmì (Clergy faction) is your
Arabic medicine consultant — you have exchanged knowledge for years.
Maimone di Palermo (People faction) supplies what cannot be found in
the apothecaries. Both know things about you the Baron must never learn.

\filetto

\subsection{The Traitor's Wife}
{\cinzel\footnotesize\color{grigionote} Costanza Ventimiglia}\\
\textit{Your husband surrendered to the Angevins. You did not.}

\textbf{Peak actions:} Resist, Deceive\\
\textbf{Secondary actions:} Plot, Persuade, Study, Manoeuvre

\textbf{Gift --- The Double Life:} In Angevin eyes, you are the
obedient wife of a collaborator. This gives you cover. Your Deceive
rolls in Angevin-controlled areas always start from Controlled position.

\textbf{Burden --- The Divided House:} Your husband does not know
what you are doing. The GM holds a secret clock: every time you act
recklessly near home, the clock advances. When it fills, he finds out.

\textbf{Opening Bond:} The Chatelaine is your cousin.
You share more secrets than is prudent.

\filetto

\subsection{The Natural Philosopher}
{\cinzel\footnotesize\color{grigionote} Greta di Caltabellotta}\\
\textit{All that moves obeys laws. All that obeys laws can be
predicted --- even men.}

\textbf{Peak actions:} Study, Persuade\\
\textbf{Secondary actions:} Plot, Command, Resist, Invoke

\textbf{Background:} Younger daughter of an inland baronial house,
Greta was given a library as her dowry instead of a convent.
She studied Latin, Greek, a few words of Arabic, and the
treatises on mechanics and astronomy her grandfather had
collected at the Hohenstaufen court. Her father tolerates her.
Her brother --- the Baron --- uses her: her mechanical
demonstrations and her tide calculations have become a social
instrument, and her evening salons in Palermo gather people who
would not otherwise speak to one another. She has no children.
She has no husbands. She has a water-driven machine that
reproduces the motion of the planets, and some people come only
to see it move.

\textbf{Gift --- The Demonstration:} When you publicly demonstrate
a principle (mechanical, mathematical, natural-philosophical),
the audience grants you authority on related matters. Once per
scene, you may use Persuade with Great effect against anyone
who has witnessed one of your demonstrations during the session.

\textbf{Burden --- The Church's Gaze:} A woman who explains how
things work attracts suspicion that men would avoid. Every time
you show off learning publicly that is not considered proper to
your station, the GM advances the Holy Office clock. When it
fills, your library is searched --- and your notes end up in the
wrong hands.

\textbf{Opening Bond:} The Chatelaine hosts your evenings. She
likes listening to you, even when she pretends otherwise. Between
you there is something neither of you has ever said.

\secsep

%% ============================================================
\section{The Clergy — Playbooks}

\subsection{The Bishop}
{\cinzel\footnotesize\color{grigionote} Berardo di Castronovo}\\
\textit{Serve God. Serve Rome. Serve your people. Choose.}

\textbf{Peak actions:} Command, Invoke\\
\textbf{Secondary actions:} Persuade, Study, Plot, Resist

\textbf{Gift --- Episcopal Authority:} Once per session, you may issue
an order that no faithful Catholic (PC or NPC) can ignore without
spiritual consequences. Those who disobey suffer 1 Stress from guilt.

\textbf{Burden --- Rome Watches:} Your movements are reported to the
curia. If you act too openly against Martin IV's interests, you risk
removal or excommunication. The GM holds a Rome clock.

\textbf{Opening Bond:} The Preaching Friar is your eyes and ears
on the street. You are not sure how much they trust you.

\filetto

\subsection{The Preaching Friar}
{\cinzel\footnotesize\color{grigionote} Fra' Tommaso di Caltagirone}\\
\textit{The word of God is a weapon. It depends only on who wields it.}

\textbf{Peak actions:} Incite, Invoke\\
\textbf{Secondary actions:} Persuade, Resist, Study, Manoeuvre

\textbf{Gift --- The Crowd:} When you address a group of people
(at least five), your Incite roll always has Standard or Great effect
— never Limited. Crowds listen to you.

\textbf{Burden --- The Fervour:} Once per session, the GM may declare
you pushed too far: the crowd reacted in a way you did not intend.
Suffer 2 Stress and manage the consequences.

\textbf{Opening Bond:} The Sacristan watches your back.
They have seen things they do not forget.

\filetto

\subsection{The Canon}
{\cinzel\footnotesize\color{grigionote} Pandolfo di Messina}\\
\textit{You know the cathedral's secrets. Including those that were never confessed.}

\textbf{Peak actions:} Study, Plot\\
\textbf{Secondary actions:} Persuade, Deceive, Resist, Invoke

\textbf{Gift --- The Archive:} Once per session, you may declare you
found information in the cathedral archives relevant to the current
situation. The GM decides what it contains, but it must be narratively
useful.

\textbf{Burden --- The Seal of Confession:} You know things you cannot
reveal. Some concern people in this story. The GM will tell you at
session start one thing you know and cannot say.

\textbf{Opening Bond:} The Bishop promoted you.
You are not certain why.

\filetto

\subsection{The Abbess}
{\cinzel\footnotesize\color{grigionote} Sister Adalasia di Troina}\\
\textit{Your convent is a refuge. For whom is up to you.}

\textbf{Peak actions:} Command, Resist\\
\textbf{Secondary actions:} Study, Persuade, Plot, Invoke

\textbf{Gift --- Sanctuary:} Your convent is sacred ground. French
soldiers cannot enter without violating canon law. You may hide
people, documents, or messages here — but everything hidden is
a responsibility.

\textbf{Burden --- My Sisters:} You have dozens of women under your
protection. Before every action that might draw Angevin attention to
the convent, roll Resist or suffer 1 Stress from the weight of
responsibility.

\textbf{Opening Bond:} The Preaching Friar brings you news from
the world. It is never good enough.

\filetto

\subsection{The Sacristan}
{\cinzel\footnotesize\color{grigionote} Arnaldo Gennarino}\\
\textit{You control the keys, the bells, the timing. More power than it seems.}

\textbf{Peak actions:} Act, Manoeuvre\\
\textbf{Secondary actions:} Study, Resist, Deceive, Invoke

\textbf{Gift --- The Bells:} You can ring the bells of Santo Spirito.
This is an act the entire city hears. Used at the right moment, it
automatically adds +1 to the faction score. Used at the wrong moment,
it may trigger an immediate Angevin response.

\textbf{Burden --- The Visibility of the Gesture:} You are the only
one who can ring those bells. If something goes wrong that evening,
everyone will know it was you.

\textbf{Opening Bond:} The Canon taught you to read.
You owe them more than that.

\filetto

\subsection{The Arab Physician}
{\cinzel\footnotesize\color{grigionote} Yusuf ibn Khalaf al-Balarmì}\\
\textit{Your family was here before the Normans. And after the Angevins.}

\textbf{Peak actions:} Study, Persuade\\
\textbf{Secondary actions:} Act, Resist, Deceive, Invoke

\textbf{Gift --- The Long Memory:} You know Palermo better than anyone
— the alleys, the passages, the wells. Once per scene, you may declare
you know an alternative route, automatically removing any physical
obstacle to movement.

\textbf{Burden --- The Stranger at Home:} You are Arab and Christian —
or Arab and still Muslim in private, decide with your GM. Either way,
you are suspect to everyone. Command rolls with people who do not
know you always start at Risky.

\textbf{Opening Bond:} The Abbess protected you once.
You do not forget.

\filetto

\subsection{The Pilgrim}
{\cinzel\footnotesize\color{grigionote} Luca di Bari}\\
\textit{You arrived in Palermo by chance. Or perhaps not.}

\textbf{Peak actions:} Manoeuvre, Deceive\\
\textbf{Secondary actions:} Study, Resist, Persuade, Invoke

\textbf{Gift --- Nobody Knows Me:} You are not on any Angevin
surveillance list. Your Manoeuvre rolls in controlled areas always
start from Controlled position.

\textbf{Burden --- Nobody Knows Me:} You have no resources, contacts,
or safe base. Every time you need logistical support, you must roll
to find it.

\textbf{Opening Bond:} The Preaching Friar fed you on your first day.
You told them your real name. No one else knows it.

\filetto

\subsection{The Copyist}
{\cinzel\footnotesize\color{grigionote} Sister Greta of Monreale}\\
\textit{Every page I copy is a choice. Every choice is a political
act --- even when no one notices.}

\textbf{Peak actions:} Study, Plot\\
\textbf{Secondary actions:} Persuade, Manoeuvre, Resist, Invoke

\textbf{Background:} Sister Greta entered Monreale at thirteen
because her mother feared she was learning too much at home.
She found the scriptorium full of Arab treatises that had
survived the searches of three regimes: mathematics, hydraulics,
automata, the gearing that drove the clocks at Roger II's court.
She copies, translates, and sometimes builds small models to
understand what the text describes. The Sacristan comes to ask
her why the Santo Spirito bell is out of tune. The Canon sends
her manuscripts he does not want to file officially. Everyone
knows that she knows. No one writes her name in the registers.

\textbf{Gift --- The Hidden Library:} You have access to texts the
official curia does not admit to possessing --- Arab engineering,
natural philosophy, forgotten records. Once per session, you may
declare you have found a text relevant to the current situation.
The GM defines the content, but it must be concrete and usable
(a mechanism, a name, a procedure).

\textbf{Burden --- Vow and Curiosity:} You took vows. They do not
conflict with your hunger to know --- but the Church does not
always agree. If you reveal to a non-religious how much you really
know, suffer 1 Stress for the breach of decorum.

\textbf{Opening Bond:} The Sacristan maintains the clocks and the
bells of Santo Spirito. You taught him how they work. He owes you
more than he knows.

\secsep

%% ============================================================
\section{The People — Playbooks}

\subsection{The Craftsman}
{\cinzel\footnotesize\color{grigionote} Calogero Ferrante}\\
\textit{You make things with your hands. You also know how to break things.}

\textbf{Peak actions:} Act, Fight\\
\textbf{Secondary actions:} Resist, Manoeuvre, Defy, Incite

\textbf{Gift --- Tools of the Trade:} You always have work tools
that can function as weapons or lockpicks without arousing suspicion.
You never have an equipment problem.

\textbf{Burden --- The Workshop:} A family depends on your work.
Every physical risk you take costs 1 Stress from the thought of them.

\textbf{Opening Bond:} The Fishwife is your neighbour.
You have had each other's backs more than once.

\filetto

\subsection{The Fishwife}
{\cinzel\footnotesize\color{grigionote} Perna di Cassaro}\\
\textit{You are at the market every morning. You know everything about everyone.}

\textbf{Peak actions:} Study, Persuade\\
\textbf{Secondary actions:} Incite, Resist, Deceive, Manoeuvre

\textbf{Gift --- Market Rumours:} Every morning at the market, you
may declare you have heard a relevant rumour. The GM tells you one
true thing about the current situation.

\textbf{Burden --- The Tongue:} You cannot stay silent when you see
injustice. Once per session, the GM may declare you said something
aloud you should not have. Manage the consequences.

\textbf{Opening Bond:} The Boy carries your baskets.
You raised them like a child. You do not want them to get hurt.

\filetto

\subsection{The Boy}
{\cinzel\footnotesize\color{grigionote} Salvo Russo}\\
\textit{You are fourteen and afraid of nothing. You should be.}

\textbf{Peak actions:} Act, Manoeuvre\\
\textbf{Secondary actions:} Defy, Resist, Incite, Study

\textbf{Gift --- Invisible:} French soldiers do not look at you —
you are just a kid. Your Manoeuvre rolls in patrolled areas always
start from Controlled position.

\textbf{Burden --- Impulsive:} When you see someone you love in
danger, roll Resist or react immediately. If you fail, you act —
and probably make things worse.

\textbf{Opening Bond:} The Fishwife raised you.
You love them like a parent. This frightens you.

\filetto

\subsection{The Fisherman}
{\cinzel\footnotesize\color{grigionote} Pietro di Cefalù}\\
\textit{You know the sea, the port, and who comes and goes. Especially who goes.}

\textbf{Peak actions:} Manoeuvre, Study\\
\textbf{Secondary actions:} Act, Resist, Fight, Deceive

\textbf{Gift --- The Port:} You have access to the dock and boats.
You can move people in or out of the city by sea without a roll,
unless the port is militarised. You also know when Charles's ships
are ready to sail.

\textbf{Burden --- The Debt:} You owe money to someone close to the
Angevins. The GM may use this debt as narrative leverage at any time.

\textbf{Opening Bond:} The Deserter (Conspirators faction) crossed
the port thanks to you. You are not yet sure if that was a mistake.

\filetto

\subsection{The Widow}
{\cinzel\footnotesize\color{grigionote} Giacoma la Nera}\\
\textit{Your husband died because of them. They did not even pay.}

\textbf{Peak actions:} Resist, Incite\\
\textbf{Secondary actions:} Persuade, Study, Fight, Invoke

\textbf{Gift --- Visible Grief:} When you speak publicly about the
wrong done to you, your Incite roll always has Great effect. Real
pain moves people.

\textbf{Burden --- The Children:} You have three small children.
They are your strength and your weakness. The GM may put one of them
in danger at any time to create narrative pressure.

\textbf{Opening Bond:} The Craftsman was your husband's friend.
He promised to keep an eye on you. You are not sure he can keep it.

\filetto

\subsection{The Street Healer}
{\cinzel\footnotesize\color{grigionote} Maimone di Palermo}\\
\textit{You are not a physician. But you know how to keep people alive.}

\textbf{Peak actions:} Act, Study\\
\textbf{Secondary actions:} Persuade, Resist, Manoeuvre, Invoke

\textbf{Gift --- Safe Hands:} When you treat someone (reduce harm
boxes), your roll always has at least Standard effect. You cannot
worsen a patient's physical condition with a failure.

\textbf{Burden --- The Vision:} You see too much. Every time you are
near someone injured, suffer 1 Stress from the weight of what you see.

\textbf{Opening Bond:} The Widow is your next-door neighbour.
Her children call you uncle.

\filetto

\subsection{The Old One}
{\cinzel\footnotesize\color{grigionote} Basilio di Monreale}\\
\textit{You have lived long enough to know how this ends. And you carry on anyway.}

\textbf{Peak actions:} Persuade, Resist\\
\textbf{Secondary actions:} Study, Incite, Invoke, Command

\textbf{Gift --- Memory:} You remember Sicily before the Angevins.
And before the Normans. Once per session, you may give another PC
+1d to their next roll by telling them how things once were.

\textbf{Burden --- The Body:} You are old. Act and Fight rolls always
start at Risky, never Controlled.

\textbf{Opening Bond:} The Boy reminds you of someone.
You do not tell them who. Perhaps that is why you love them.

\filetto

\subsection{The Workshop Mistress}
{\cinzel\footnotesize\color{grigionote} Greta the Tinsmith}\\
\textit{People bring you broken things. You give them back healed.
While they wait, they tell you everything.}

\textbf{Peak actions:} Persuade, Act\\
\textbf{Secondary actions:} Study, Incite, Manoeuvre, Resist

\textbf{Background:} Greta's bench stands at the corner of
Seralcadio, where the fishermen pass on their way down to the
port and the artisans on their way up from the Cassaro. She fixes
locks, repairs buckets and scales, solders pans, mends small
clocks when they come her way. She shows the children a little
automaton of her own --- a dove that beats its wings when you
turn a crank --- and for that reason half the quarter knows her
name. Her father was a tinsmith and she took the bench at sixteen.
Since then the bench has become a tribunal, a meeting place, a
confessional, an information base. She mends. People talk.

\textbf{Gift --- The Open Bench:} Your workshop is a public square.
Once per scene, you may declare that a specific person (PC or
NPC currently in the area) has come to you --- to fix something,
to watch, or just to talk. From that moment they are available
for conversation, in narratively neutral position.

\textbf{Burden --- Everyone Sees You:} Your visibility is your
strength and your trap. Angevin officials know your name. Every
time you organise something concrete in the quarter (meetings,
signals, distributions), the GM advances the ``soldiers at your
door'' clock. When it fills, they come.

\textbf{Opening Bond:} The Fishwife brings you customers. In return
you adjust the weights on her scales --- and she knows you know.
When the moment comes, you will discover who is protecting whom.

\filettodoppio
